Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-26 RevisedSubmission/hines_hetland_2017_gji.tex
did more

We consider the noise vector, $\mitbf{\eta}_i$, to be composed of a
temporally correlated Gaussian process, $\mathcal{L}_i \sim
\mathcal{GP}(0,C_{\mathcal{L}_i})$, and a vector of uncorrelated
Gaussian noise, $\mitbf{w}_i$, so that
\begin{equation}
\mitbf{\eta}_i = \mathcal{L}_i(\points) + \mitbf{w}_i.
\end{equation}
The standard deviations for $\mitbf{w}_i$ are given by the
uncertainties derived for the GNSS displacement solutions,
$\mitbf{\sigma}_i$. The noise vector then has zero mean and the
covariance matrix
\begin{equation}
\mitbf{C}_{\eta_i} = C_{\mathcal{L}_i}(\points,\points) +
                     \mathrm{diag}(\mitbf{\sigma}_i^2).
\end{equation}

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades \citep[e.g.,][]{Zhang1997, Mao1999,
Williams2004, Langbein2008}. In these studies, temporally correlated
noise tends to be described with some combination of Brownian motion
(also known as random walk noise or a Weiner process), a first-order
Gauss-Markov (FOGM) process, and/or flicker noise. There is some
physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. In particular, the power spectrums for Brownian motion and
monument motion are both proportional to $f^{-2}$, where $f$ is
frequency. Brownian motion, by definition, contains a reference time
at which the process begins. Since there is no notion of when noise
``begins" in GNSS data, we do not find Brownian motion to be an
appropriate noise model. On the other hand, a FOGM process also has a
power spectrum that is proportional to $f^{-2}$, above some corner
frequency, and it has no reference time (i.e., a FOGM process is
stationary) \citep{Langbein2004}. We then let $\mathcal{L}_i$ be a
spatially uncorrelated FOGM process, which has the covariance function
\begin{equation}\label{eq:FOGM}
C_{\mathcal{L}_i}\left((\pos,t),(\pos',t')\right) = 
\frac{\beta^2}{2\alpha}
\exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'}
\end{equation}
\citep[sec. B2.1]{Rasmussen2006}, where $\delta_{\pos,\pos'}$ is 1 if
$\pos = \pos'$ and 0 otherwise.

@@

We constrain the parameters for $\mitbf{\eta}_i$, $\alpha$ and
$\beta$, with a set of 38 continuous GNSS stations in the Pacific
Northwest that are east of 121$^\circ$W.  These stations are
sufficiently far from the subduction zone that they are unlikely to
contain transient signal associated with SSEs. We find $\alpha$ and
$\beta$ for each station time series with the Restricted Maximum
Likelihood (REML) method \cite[e.g.,][]{Harville1974, Cressie1992}.
The REML method finds the hyperparameters, which we collectively refer
to as $\mitbf{\theta}$, that maximize the likelihood function
\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_*^T\mitbf{K}(\mitbf{\theta})\data_*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{\Sigma}(\mitbf{\theta})^{-1}.
\end{equation}
\citet{Harville1974} showed that choosing the hyperparameters which
maximize eq. (\ref{eq:REML}) is equivalent to choosing the
hyperparameters which maximize the probability of drawing $\data_*$
from $\data$. We use the REML method over the maximum likelihood
method \citep[e.g.,][]{Langbein1997} because the REML method accounts
for the improper prior that we assigned to $\mitbf{a}$
\citep{Hines2017}. We independently estimate $\mitbf{\theta}$ for each
station, and so $\data_*$ consists of displacements for an individual
station. We are assuming $u(p)=0$ when estimating the noise
hyperparameters for this section.

@@

The distribution of inferred $\alpha$ and $\beta$ are shown in Figure
\ref{fig:NoiseParams}. The amplitude of FOGM noise, $\beta$, for the
easting and northing components is notably low and are clustered
around 0.5 mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend
to cluster around 0 yr$^{-1}$, suggesting that noise can be described
with Brownian motion. We also estimate hyperparameters for the
vertical component of displacements, under the hope that vertical
deformation gradients could reveal some geophysical signal. The
amplitude of FOGM noise for the vertical component is relatively large
with a median value of 13.5 mm/yr$^{0.5}$.  The inferred values for
$\alpha$ are also higher for the vertical component with a median
value of 8.21 yr$^{-1}$. In Figure \ref{fig:NoiseSamples}, we use the
median values of $\alpha$ and $\beta$ to generate two random samples
of FOGM noise for each component. The samples span five years and over
these five years the easting and northing samples drift by about 1 mm.
In the context of detecting SSEs, which produce several mm's of
surface displacement on the time-scale of weeks, the estimated FOGM
noise for the easting and northing component is negligible. In
contrast, the estimated FOGM noise for the vertical component is
larger than the signal we would expect from SSEs. We suspect that the
higher amplitude for the FOGM noise in the vertical component is
accommodating for deficiencies in our rather simple seasonal model.
Based on this analysis, we henceforth ignore temporally correlated
noise in the easting and northing component because of its low
amplitude. We also do not use vertical displacements because of the
presumably low signal-to-noise ratio.

\begin{figure*}
\includegraphics{figures/noise/noise-params.pdf}
\caption{
Distribution of estimated FOGM hyperparameters (eq. \ref{eq:FOGM}).
Hyperparameters are estimated with the REML method for 38 stations in
the Pacific Northwest that are east of $121^\circ$W. ``IQR'' is the
interquartile range.
}   
\label{fig:NoiseParams}
\end{figure*}

\begin{figure*}
\includegraphics{figures/noise/noise-samples.pdf}
\caption{
Two FOGM noise samples for each component. The FOGM hyperparameters
have been set to the median values from Figure \ref{fig:NoiseParams}.
}   
\label{fig:NoiseSamples}
\end{figure*}

@@

Another significant source of noise in GNSS data is common mode error
\citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly
spatially correlated. When not accounted for, common mode error
manifests as spatially uniform undulations in estimated transient
displacements. However, estimated transient strain rates are
insensitive to common mode error. We therefore do not include common
mode error in our noise model. We then make the simplifying assumption
that $\eta(p) = 0$ for the easting and northing component of GNSS
data.

>>>>

We consider the noise vector, $\mitbf{\eta}_i$, to be composed of a
temporally correlated Gaussian process, $\mathcal{L}_i \sim
\mathcal{GP}(0,C_{\mathcal{L}_i})$, and a vector of uncorrelated
Gaussian noise, $\mitbf{w}_i$, so that
\begin{equation}
\mitbf{\eta}_i = \mathcal{L}_i(\points) + \mitbf{w}_i.
\end{equation}
The standard deviations for $\mitbf{w}_i$ are taken to be the
uncertainties derived for the GNSS displacement solutions,
$\mitbf{\sigma}_i$. The noise vector then has zero mean and the
covariance matrix
\begin{equation}
\mitbf{C}_{\eta_i} = C_{\mathcal{L}_i}(\points,\points) +
                     \mathrm{diag}(\mitbf{\sigma}_i^2).
\end{equation}

% Justifying our temporal noise model

The temporally correlated noise in GNSS data has been thoroughly
studied over the past two decades \citep[e.g.,][]{Zhang1997, Mao1999,
Williams2004, Langbein2008}. In these studies, temporally correlated
noise tends to be described with some combination of Brownian motion
(also known as random walk noise or a Weiner process), a first-order
Gauss-Markov (FOGM) process, and/or flicker noise. There is some
physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. In particular, the power spectrums for Brownian motion and
monument motion are both proportional to $f^{-2}$, where $f$ is
frequency. Brownian motion, by definition, contains a reference time
at which the process begins. Since there is no notion of when noise
``begins" in GNSS data, we do not find Brownian motion to be an
appropriate noise model. On the other hand, a FOGM process also has a
power spectrum that is proportional to $f^{-2}$, above some corner
frequency, and it has no reference time (i.e., a FOGM process is
stationary). We then let $\mathcal{L}_i$ be a spatially uncorrelated
FOGM process, which has the covariance function
\begin{equation}\label{eq:FOGM}
C_{\mathcal{L}_i}\left((\pos,t),(\pos',t')\right) = 
\frac{\beta^2}{2\alpha}
\exp\left(-\alpha|t - t'|\right)\delta_{\pos,\pos'}
\end{equation}
\citep[sec. B2.1]{Rasmussen2006}, where $\delta_{\pos,\pos'}$ is 1 if
$\pos = \pos'$ and 0 otherwise.

@@

By choosing a FOGM process for $\mathcal{L}_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
where we pick $\mitbf{\theta}$ to maximize the likelihood of drawing
the observed data, $\mitbf{d}_i^*$, from the random vector
$\mitbf{d}_i$. However, unlike the MLE method, the REML method
produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to independently estimate
$\mitbf{\theta}$ at 38 continuous GNSS stations in the Pacific
Northwest that are east of 121$^\circ$W. These stations are
sufficiently far from the subduction zone that they are unlikely to
contain transient signal associated with SSEs. We assume that the
noise at these inland stations is representative of the noise at all
the stations considered in this study, which is probably a poor
assumption since the inland stations are subject to distinctly
different climatic conditions. Nonetheless, we find the
$\mitbf{\theta}$ for each inland station and for each displacement
component that maximizes the REML likelihood function
\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*},
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}

@@

The distribution of inferred $\alpha$ and $\beta$ are shown in Figure
\ref{fig:NoiseParams}. The estimates of $\beta$ for the easting and
northing components are notably low and are clustered around 0.5
mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster
around 0 yr$^{-1}$. We also estimate $\alpha$ and $\beta$ for the vertical
component of displacements, with the hope that vertical deformation
gradients could reveal some geophysical signal. For the vertical
component, $\beta$ is relatively large with a median value of 13.5
mm/yr$^{0.5}$. The inferred values for $\alpha$ are also higher for
the vertical component with a median value of 8.21 yr$^{-1}$. In
Figure \ref{fig:NoiseSamples}, we use the median values of $\alpha$
and $\beta$ to generate two realizations of FOGM noise for each
component. The realizations span five years, and the easting and
northing realizations drift by about 1 mm over these five years. In
the context of detecting SSEs, which produce several mm's of surface
displacement on the time-scale of weeks, the estimated FOGM noise for
the easting and northing component is negligible. In contrast, the
estimated FOGM noise for the vertical component is larger than the
signal we would expect from SSEs. We suspect that the higher amplitude
for the FOGM noise in the vertical component is accommodating for
deficiencies in our rather simple seasonal model. Based on this
analysis, we henceforth ignore temporally correlated noise in the
easting and northing component because of its low amplitude. We also
do not use vertical displacements because of the presumably low
signal-to-noise ratio.

\begin{figure*}
\includegraphics{figures/noise/noise-params.pdf}
\caption{
Distribution of estimated FOGM parameters (eq. \ref{eq:FOGM}).
parameters are estimated with the REML method for 38 stations in
the Pacific Northwest that are east of $121^\circ$W. ``IQR'' is the
interquartile range.
}   
\label{fig:NoiseParams}
\end{figure*}

\begin{figure*}
\includegraphics{figures/noise/noise-samples.pdf}
\caption{
Two FOGM noise samples for each component. The FOGM parameters
have been set to the median values from Figure \ref{fig:NoiseParams}.
}   
\label{fig:NoiseSamples}
\end{figure*}

@@

Another significant source of noise in GNSS data is common mode error
\citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly
spatially correlated. When not accounted for, common mode error
manifests as spatially uniform undulations in estimated transient
displacements. However, estimated transient strain rates are
insensitive to common mode error, and so we do not need to include
common mode error in our noise model.
